---
title: "scRNA_Day2"
author: "Roxanne Kahn-Cleland"
date: "2024-06-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

install.packages("viridis", repos = "http://cran.us.r-project.org")
```

# Annotate cell types

Step.1 Load Library and your seurat object

```{r}
library(SingleR)
library(Seurat)
library(celldex)
library(dplyr)
library(pheatmap)

sobj <- readRDS("Output_ovarian_carcinoma.rds") 
# our output file from initial scRNA analysis part 1
sobj
```

## Step 2. Prepare Inputs for running automatic cell type annotation

First we need to extract the normalized data using GetAssayData().

```{r}
sobj_count <- GetAssayData(sobj, layer = "data") 
# 3 layers present: counts, data, scale.data (from table before)
str(sobj_count)
```

```{r}
ref <- celldex::BlueprintEncodeData()
# celldex:: to show where we took ref from
# set up reference scRNA annotation
ref # read/show reference annotation 

```

```{r}
library(SingleR)
pred <-SingleR(test=sobj_count, ref=ref, labels = ref$label.main)
pred %>%  head() # add the predicted cell types to our existing file = column labels
```

```{r cars}
sobj$labels <- pred[match(rownames(sobj@meta.data), rownames(pred)), 'labels']
# tries to match row  names from seurat object to prediction
# labels hidden in table underneath because too small, us the arrow to go to the right
# links cell clusters to predictions for cell types
sobj@meta.data %>% head()
```

## Step 4. Visualize the data

This time we will group by "Labels", since that is the column that contains the predicted cell type.

```{r pressure, echo=FALSE}
umap_pred <-DimPlot(
  sobj, reduction = "umap", group.by = "labels", label = TRUE)
 umap_pred
 # we are grouping by the column labels i.e the predcited cell types, but label = TRUE gives us labels on the graph
```

Compare with previous data.

```{r}
umap_sobj <-DimPlot(
  sobj, reduction = "umap", group.by = "seurat_clusters", label = TRUE)

umap_sobj | umap_pred
# predictions and clusters dont exactly overlap, do we need to make subsets?
# is astrocyte and c´the green cluster correct? look in heatmap
```

Plot the Single R assignment score across all cell.label combination

```{r}
#install.packages("viridis")
plotScoreHeatmap(pred)
# labels left are cell types from the reference
# labels on right = cells from our dataset
# look at colour bar up top, then go down the heatmap and find the areas that are brightest yellor, this are the cells that most fit

```

## Step 2.

We just compared the reference, now we should also do the marker based method for a robust prediction of cell type.

Marker-based: cellMarker 2.0

```{r}
human_marker <- read.csv("CellMakerv2_Human.csv")
str(human_marker)
# here we get markers to compare our own dataset
```

Select marker for ovary from the tissue class column.

```{r}
ovary_marker <- human_marker %>% 
  filter(species=="Human" & tissue_class=="Ovary") %>% 
  select(c(3,5,6,7,8,9,16)) 
# filter for only markers we ant to look for
# select the columns we want from human_marker file
# while filtering for human and ovary

ovary_marker %>% head()
```

Get markers table from your seurat object:

```{r}
sobj.markers <- FindAllMarkers(sobj, only.pos = TRUE)
sobj.markers %>% head()

# p value significance
# how higher or lower is expression compared to rest of clusters
# pct 1, percentage of cells in tis cluster that expressed this gene
# pct 2, how many cells of the other clusters expressed this gene
# p vald adjustment = how high is false discovery rate

```

Combine CellMarker with our markers sobj.markers

```{r}
merged_cellmarker <- sobj.markers %>% filter(p_val_adj<0.05 & avg_log2FC > 1 & pct.1 > 0.8 & pct.2 < 0.5) %>% 
  left_join(ovary_marker[,c(3,4,6)], join_by(gene==marker), relationship = "many-to-many")

# we only accept false discover under 0.05 = p_val_adj
# we want gene to be expressed in at least 80% of cell cluster, but also under 50% in other clusters
# log_fold change = everything over 1 = we want at least a 2 fold log change higher in expression
# relationship = parameter within left join, we try to match our marker table with ovary marker file, will probably get a warning because multiple will match = so we add "many-to-many relationship to allow multiple matching
# if we seöect PAX as a amrker for ovariancells, but PAX is also expressed in other clusters, if we dont use the many.ti many option, it will only turn up the first match instead of all

# if look at merged_cell = some cells are defined, yet many N.a because not all cell markers are defined?

# merged our markers and referene markers?

merged_cellmarker %>% head()
```

Let's try to see if there are fibroblasts and astrocytes:

```{r}
merged_cellmarker %>% 
  filter(cluster==2 & cell_type=="Cancer cell") %>% 
  select(c(7,8,9))
# merged our markers and referene markers?
```

From the results, we found PDGFRA, CAV1, ACTA2 pointing to fibroblast in cancer cells

check again using FeaturePlot()

```{r}
#genes <- c("PDGFRA", "CAV1", "ACTA2") #i made a mistake at the very beginning of data analysis, we didnt find these genes
# correct code

genes <- c("ALDH1A1")
FeaturePlot(sobj, features = genes)
```

## Step 6. Annotate the cells

Rename cluster 2 to Fibroblasts using RenameIdents()

```{r}
sobj_new <- RenameIdents(sobj, "2" = "Fibroblasts")
# rename the cluster 2 = Fibroblast then use Dimplot to plot it
DimPlot(sobj_new, reduction ="umap", label = T)
```

## Step 7. Differentially expressed genes

Can we distinguish cluster 7?

```{r}
merged_cellmarker %>% 
  filter(cluster==7 & cell_type=="Cancer cell") %>% 
  select(c(7,8,9)) %>%  # only need last 3 columns
  str()

# here we get choice of mysenchymal cells and fibroblast (in yaos version without mistake)
# in reference method in coloured plot we saw that they may be adipocytes

```

```{r}
BiocManager::install("clusterProfiler") # for checking function of subset of genes = ontology
BiocManager::install("org.Hs.eg.db") # human database as example
BiocManager::install("DOSE") # dot plot visualizer

library(org.HS.eg.db)
library(clusterProfiler)
library(DOSE)
```

```{r}
# proper code
# first we ask whether is mysenchymal or fibroblast
#marker_7_2 <- FindMarkers(sobj, ident.1 = 7, ident.2 =2)
#marker_7_2_sig <- marker_7_2 %>% 
#  filter(abs(avg_log2FC)>= 1 & p_val_adj<0.05 & pct.2<0.4)

marker_3_2 <- FindMarkers(sobj, ident.1 = 3, ident.2 =2)
marker_3_2_sig <- marker_3_2 %>% 
  filter(abs(avg_log2FC)>= 1 & p_val_adj<0.05 & pct.2<0.4)
str(marker_3_2)
```

```{r}
log2 <- marker_3_2_sig$avg_log2FC
names(log2) <- rownames(marker_3_2_sig)

log2_sort <- sort(log2, decreasing = TRUE)
log2_sort[1:5]

#we select average_log2FC => differentially expressed vs others?
```

Run gene set enrichment analysis:

```{r}
gse <- gseGO(geneList = log2_sort,
             ont="ALL",
             keyType="SYMBOL", # if ensemble or NCBI ID must specify here
             nPerm=10000, #larger number of permutations give you a more precise result
             OrgDb = "org.Hs.eg.db") # database we want to map = here human = Hs = homo sapiens


#take subset of genes and compare to reference
# log2_sort is expression difference between cluster 3 and 2 (7 and 2)
# ont could be BP etc....
# name of gene is gene symbol, 
# interested? go learn more about GSEA


gse <- gseGO(geneList = log2_sort, 
             ont="ALL",
             keyType="SYMBOL",
             nPerm=100000,
             OrgDb= "org.Hs.eg.db")
dotplot(gse)
```

Visualize the Results:

```{r}
enrichplot::dotplot(gse)
# doesnt work for me, need to use leas data file from friday
# my warning says subset didnt match
library("stats")

# fibroblast or mysenchymal cells?
# 700 genes that differ between cluster 7 and 2
# we know cluster 2 is fibrolblast so probably 7 ist fibro
# we check the plot here and see the cells play a big role in immune response
# cluster 7 more likely to be mesenchymal than adipocytes, because play bigger role in immune response? this is a guess

# we know cluster 2 is fibroblast => single R calculates correlation between dataset and celldex blueprint refernce transcriptome => get a distribution that tells us which one is most likely
# single R makes a prediction, we put this prediction back into our seurat object
# in plot we saw cluster 7 ist astrovcyte/fibroblast. Which one is which?
# we then use the cell markers from Cellmarker, get sonme refernce markers for ovarian cancer = now we generate markers fron our seurat object so that we can compare markers
# set a strict threshold for a robust answer + select for ovary cell => merged_cellmarker
# most genes are N.a because not specific to the ovary
# now we select for cancer cell = get a fibroblast result

```

Put the identity back

```{r}
sobj_new <- RenameIdents(sobj, "3" = "Mesenchymal Cells",
                         "2" = "Fibroblasts") 
# 7 in normal code = mesenchymal cells
DimPlot(sobj_new, reduction = "umap", label = TRUE)
```

Upload to gitlab

```{r}
# go to git-lab
# creat a project: scRNAseq_turtorial
# now go back to R
# click the R+ signh up top to create R project (public and in this case we did a README)
# select version control => git
# got to gut lab => click code button, copy the ssh link
# use that link as a repository 
# browse and save folder onto desktop
# put the rmarkdown file/html and put it into that folder on desktop 
# can write into file, there should be a new git button in the environment pane menu => there is also a commit button, can then also write a commit message, then click commit button
# then press push button
```
